RegressionPvNP

### 2017-05-17 Clustering.Rmd
…

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(plyr)
library(xlsx)
library(dplyr)
library(pander)
library(Hmisc)
library(ggdendro)
library(reshape2)
library(GenomicRanges)
library(gridExtra)
library(biomaRt)
library(multtest)
library(tibble)
library(mclust)
library(ggfortify)

@@

## Patient info file
patient.file = grep('All_patient_info.xls', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

load('patients.Rdata', verbose=T)

@@

```{r pca, echo=F, warning=F, message=F, fig.width=8, fig.height=4}
sum.patient.data = arrange(sum.patient.data, Status, total.samples)

plot.pca<-function(pt) {
  print(pt)
  info = arrange(patient.data[[pt]]$info, Endoscopy.Year)
  info = info[which(info$Samplename %in% colnames(patient.data[[pt]][['seg.vals']])[-(1:5)]),]
  plots = list()
  for (sv in c('seg.vals')) {

    x1 = patient.data[[pt]][[sv]][, info$Samplename]
    x1 = x1[,info$Samplename]
    colnames(x1) = paste(info$Endoscopy.Year, '(', info$Pathology, ')', sep='' )

    pca = prcomp(t(x1), scale=T)

@@

```{r echo=F, message=F, warning=F, pcaplots, fig.align='left', fig.height=30, fig.width=20}
grid.arrange(grobs=pplots, nrow=length(subset(sum.patient.data, Status == 'P')$Patient), top="Progressors")
#grid.arrange(grobs=npplots, nrow=length(subset(sum.patient.data, Status == 'NP')$Patient), top="Non-Progressors")
```

>>>>

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(plyr)
library(dplyr)
library(pander)
library(Hmisc)
library(ggdendro)
library(reshape2)
library(GenomicRanges)
library(gridExtra)
library(biomaRt)
library(multtest)
library(tibble)
library(mclust)
library(ggfortify)

@@

## Patient info file
patient.file = grep('All_patient_info.txt', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

load('Training_patients.Rdata', verbose=T)

@@

```{r pca, echo=F, warning=F, message=F, fig.width=8, fig.height=4, eval=F}
sum.patient.data = arrange(sum.patient.data, Status, total.samples)

plot.pca<-function(pt) {
  print(pt)
  info = arrange(patient.data[[pt]]$info, Endoscopy.Year)
  if (length(info$Samplename) <= 1) return(NA)

  info = info[which(info$Samplename %in% colnames(patient.data[[pt]][['seg.vals']])[-(1:5)]),]
  plots = list()
  for (sv in c('seg.vals')) {

    x1 = patient.data[[pt]][[sv]][, info$Samplename]

    if (length(info$Samplename) <= 1) { 
      x1 = as.data.frame(x1)
      colnames(x1) = info$Samplename
    } else {
      x1 = x1[, info$Samplename]
    }

    colnames(x1) = paste(info$Endoscopy.Year, '(', info$Pathology, ')', sep='' )

    #pca = prcomp(t(x1), scale=length(info$Samplename) > 2)
    pca = prcomp(t(x1), scale=length(info$Samplename) > 2)

@@

```{r echo=F, message=F, warning=F, pcaplots, fig.align='left', fig.height=30, fig.width=20, eval=F}
grid.arrange(grobs=pplots, nrow=length(subset(sum.patient.data, Status == 'P')$Patient), top="Progressors")
#grid.arrange(grobs=npplots, nrow=length(subset(sum.patient.data, Status == 'NP')$Patient), top="Non-Progressors")
```

### 2017-05-17 Patient Report.Rmd
…

## Patient info file
patient.file = grep('/All_patient_info.xls', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

dataset = 'Training'
#dataset = 'Test'
patient.info = read.patient.info(patient.file, set=dataset)
patient.info$Patient = gsub("/", "_", patient.info$Patient)
head(patient.info)

bad_batch = which(patient.info$Batch.Name %in% c('Batch_8', 'Batch_9'))
patient.info = patient.info[-(bad_batch),]

@@

lowprobes = c()

patient.data = list()

@@

  if(!file.exists(paste(patient.plot.dir, "/",patient.name,"_segmentedCoverage_fitted_gamma",gamma2,".txt",sep=""))) next

  segvals = read.table(paste(patient.plot.dir, "/",patient.name,"_segmentedCoverage_fitted_gamma",gamma2,".txt",sep=""),sep="\t",stringsAsFactors=F,header=T)

@@

  if (length(missing) > 0) {
    warning(paste("Missing", length(missing), "sample(s) in patient", patient.name))
    next
  }

  probes = length(which(segvals$n.probes < min.probes))
  message( paste(probes, ' probes (', round(probes/nrow(segvals))*100, '%)',' below the minimum probe count (',min.probes,')', sep=''))
  if (probes/nrow(segvals) >= 0.75) lowprobes = c(lowprobes, patient.name)

@@

  if (nrow(segvals) < 100) {
    warning(paste(patient.name, 'has fewer than 100 genomic reginos with segmented values:', nrow(segvals)))
    next
  }

@@

  sample.cols = grep('^D\\d+_D\\d+', colnames(segvals))
  if (nrow(segvals) > 0) {
      HC = hclust(dist(t(segvals[,sample.cols])))
      gg1 = ggdendrogram(HC, rotate=T) + labs(title = paste(patient.name,": fitted segment coverage"),x="")
      #print(gg1)

      # Normalize  (value-mean(value))/sd(value)
      normalised.segvals = segvals[,sample.cols]
      for(c in 1:nrow(normalised.segvals)) 
        normalised.segvals[c,] = (normalised.segvals[c,]-mean(unlist(normalised.segvals[c,])))/sd(unlist(normalised.segvals[c,]))

      patient.data[[patient.name]][['norm.seg.vals']] = normalised.segvals

      # Variable Regions
      sds = apply(segvals[,sample.cols],1,sd)
      means = apply(segvals[,sample.cols],1,mean)
      CofV = sds/means

      variable.region.indices = which(sds >= sd.threshold & segvals$n.probes >= min.probes)
      variable.regions = segvals[variable.region.indices, intersect(patient.data[[patient.name]]$info$Samplename, colnames(segvals))] 
      patient.data[[patient.name]][['variable.regions']] = cbind(segvals[variable.region.indices,c(1:5)],variable.regions )

@@

      gg3 = ggplot( melt(normalised.segvals), aes(x=value) ) + 
        facet_wrap(~variable,scales = "free_x") + 
        geom_histogram( fill=I("blue"),col=I("lightblue"),binwidth=0.3) + labs(title=paste(patient.name,'normalised segment coverage'), x='', y='')

      #if(nrow(segvals)>=3) {
        # order by date of endoscopy
        patient = subset(patient.info, Patient == patient.name)
        patient = arrange(patient, Endoscopy.Year, Pathology, Samplename)
       # }
        sample.normalised.segvals = normalised.segvals[, intersect(colnames(normalised.segvals), patient$Samplename)]
      }
}

print(paste("Missing patients:", paste(setdiff(sum.patient.data$Patient, names(patient.data)), collapse=', ')))

@@

save(patient.data, file=paste(dataset, '_patients.Rdata', sep=''))

@@

chrs = read.table('hg19_chrom.txt',sep='\t', header=T)

# Segment lengths, variable regions
var.seg.lengths = lapply(var.seg.lengths, function(x) merge(x, chrs, by.x='chr', by.y='chr') )

# All segments
segment.lengths = lapply(segment.lengths, function(x) merge(x, chrs, by.x='chr', by.y='chr') )

# Genome coverage
genome.coverage = sapply(segment.lengths, function(x) {
  sum(as.numeric(x[['seg.len']]))/sum(as.numeric(chrs$length))
})

@@

var.seg.lengths = lapply(var.seg.lengths, function(df) {
  as.data.frame(df %>%
  group_by(chr, length) %>%
  summarise(seg.lengths = sum(seg.len), 
            chr.ratio = sum(seg.len)/length[1]))
})

@@

rm(var.regions)
for (patient.name in names(patient.data)) {
  regions = patient.data[[patient.name]]$variable.regions[,c('chrom','arm','start.pos','end.pos')]
  if (nrow(regions) <= 0) next

@@

gene.cnt = var.regions %>%
  group_by(Gene.Symbols, CN) %>%
  summarise(Progressors=length(which(status == 'P')), Non.Progressors=length(which(status == 'NP')) )
gene.cnt[,unique(var.regions$patient.name)] = 0
gene.cnt = gene.cnt[which(!is.na(gene.cnt$Gene.Symbols)),]

@@

m = melt(gene.p[,c(1:4)], id.vars=c('Gene.Symbols', 'CN'))
m$Gene.Symbols = factor(m$Gene.Symbols, levels= m$Gene.Symbols[c(nrow(m):1)] ) # order
m$CN = factor(m$CN)

ggplot(m, aes(y=Gene.Symbols, x=variable)) + 
  geom_tile(aes(fill=value), color='white') + 
  scale_fill_gradient('n patients',low='white',high='red') + labs(x="",y='CGC Genes')
# + geom_point(data=m[m$CN == 'gain' & m$value > 0,], aes(y=Gene.Symbols, x=variable), size=2, shape=24, fill='red', color='red') +
#  geom_point(data=m[m$CN == 'loss' & m$value > 0,], aes(y=Gene.Symbols, x=variable), color='green4', size=2, shape=25, fill='green4')

gc = gene.cnt %>% group_by(Gene.Symbols) %>% summarise(Progressors=sum(Progressors), Non.Progressors=sum(Non.Progressors))
gc = arrange(subset(gc, Progressors >= Non.Progressors), -Progressors)
gc = gc[1:60,]
m = melt(gc, id.vars=c('Gene.Symbols'))
#m$value = as.factor(m$value)
m$Gene.Symbols = factor(m$Gene.Symbols, levels= m$Gene.Symbols[c(nrow(m):1)] ) # order
ggplot(m, aes(y=Gene.Symbols, x=variable)) + 
  geom_tile(aes(fill=value), color='white') + 
  scale_fill_gradient('n patients',low='white',high='red') + labs(x="",y='CGC Genes') + ggtitle("Top 50 genes only")

@@

pvnp = data.frame(matrix(nrow=length(subset(sum.patient.data, Status == 'P')$Patient), ncol=2, dimnames=list(subset(sum.patient.data, Status == 'P')$Patient, c('hgd','be'))))    
for (name in subset(sum.patient.data, Status == 'P')$Patient) {
  sn = which(colnames(patient.data[[name]]$variable.regions)[-(1:5)] %in% subset(patient.data[[name]]$info, grepl('HGD|IMC', Pathology))$Samplename)
  if (length(sn) <= 0 | nrow(patient.data[[name]]$variable.regions) <= 0) next

@@

This is just looking at the distribution of values per sample. The average ploidy per sample is around 1.  Currently we're calling variable regions based on this.
```{r, warning=F}
ks = list()
for (pt in names(patient.data)) {
  cols = intersect(arrange(patient.data[[pt]]$info, Pathology, Endoscopy.Year)$Samplename, colnames(patient.data[[pt]]$seg.vals[,-(1:5)]))
  segvals = patient.data[[pt]]$seg.vals[,cols]

>>>>

## Patient info file
patient.file = grep('/All_patient_info.txt', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

dataset = 'Training'
#dataset = 'Test'
patient.info = read.patient.info(patient.file, set=dataset)
head(patient.info)

@@

cachefile = paste(dataset, '_patients.Rdata', sep='')
if ( file.exists(cachefile) ) {
  load(cachefile, verbose=T)
} else {
  patient.data = list(); lowprobes = c()

@@

    if (!file.exists(paste(patient.plot.dir, "/",patient.name,"_segmentedCoverage_fitted_gamma",gamma2,".txt",sep=""))) {
      warning(paste("Missing segmented file for", patient.name))
      next
    }

    segvals = read.table(paste(patient.plot.dir, "/",patient.name,"_segmentedCoverage_fitted_gamma",gamma2,".txt",sep=""),sep="\t",stringsAsFactors=F,header=T)
    dim(segvals)

@@

    if (length(missing) > 0) {
      warning(paste("Missing", length(missing), "sample(s) of", length(subset(patient.info, Patient == patient.name)$Samplename), "in patient", patient.name))
      #next
    }

    probes = length(which(segvals$n.probes < min.probes))
    message( paste(probes, ' probes (', round(probes/nrow(segvals),2)*100, '%)',' below the minimum probe count (',min.probes,')', sep=''))
    if (probes/nrow(segvals) >= 0.75) lowprobes = c(lowprobes, patient.name)

@@

    if (nrow(segvals) < 100) {
      warning(paste(patient.name, 'has fewer than 100 genomic regions with segmented values:', nrow(segvals)))
      #next
    }

@@

    sample.cols = grep('^D\\d+_D\\d+', colnames(segvals))

    if (length(sample.cols) > 0) {
      if (length(sample.cols) > 2) {  
        HC = hclust(dist(t(segvals[,sample.cols])))
        gg1 = ggdendrogram(HC, rotate=T) + labs(title = paste(patient.name,": fitted segment coverage"),x="")
        #print(gg1)
      }

      # Normalize  (value-mean(value))/sd(value)
      normalised.segvals = segvals[,sample.cols]
      if (length(sample.cols) > 1) {  
        for(c in 1:nrow(normalised.segvals)) 
          normalised.segvals[c,] = (normalised.segvals[c,]-mean(unlist(normalised.segvals[c,])))/sd(unlist(normalised.segvals[c,]))
        patient.data[[patient.name]][['norm.seg.vals']] = normalised.segvals

@@

        gg3 = ggplot( melt(normalised.segvals), aes(x=value) ) + 
          facet_wrap(~variable,scales = "free_x") + 
          geom_histogram( fill=I("blue"),col=I("lightblue"),binwidth=0.3) + labs(title=paste(patient.name,'normalised segment coverage'), x='', y='')
      }

      # Variable Regions can only be called relative to each sample so if there's only one sample we can't make a call
      if (length(sample.cols) > 1) {
        sds = apply(segvals[,sample.cols],1,sd)
        means = apply(segvals[,sample.cols],1,mean)
        CofV = sds/means

        variable.region.indices = which(sds >= sd.threshold & segvals$n.probes >= min.probes)
        variable.regions = segvals[variable.region.indices, intersect(patient.data[[patient.name]]$info$Samplename, colnames(segvals))] 
        patient.data[[patient.name]][['variable.regions']] = cbind(segvals[variable.region.indices,c(1:5)],variable.regions )
      }

      patient = arrange(subset(patient.info, Patient == patient.name),  Endoscopy.Year, Pathology, Samplename)
    }
  }
save(patient.data, file=paste(dataset, '_patients.Rdata', sep=''))
}
print(paste("Missing patients:", paste(setdiff(sum.patient.data$Patient, names(patient.data)), collapse=', ')))

@@

@@

chr.info = read.table('hg19_chrom.txt',sep='\t', header=T)

# Segment lengths, variable regions
var.seg.lengths = lapply(var.seg.lengths, function(x) merge(x, chr.info, by.x='chr', by.y='chr') )

# All segments
segment.lengths = lapply(segment.lengths, function(x) merge(x, chr.info, by.x='chr', by.y='chr') )

# Genome coverage
genome.coverage = sapply(segment.lengths, function(x) {
  sum(as.numeric(x[['seg.len']]))/sum(as.numeric(chr.info$length))
})

@@

var.seg.lengths =  lapply(var.seg.lengths, function(df) {
   ddply(df, .(chr, length), summarise, seg.lengths = sum(seg.len), chr.ratio = sum(seg.len)/length[1])
})

@@

rm(var.regions)
for (patient.name in names(patient.data)) {
  regions = patient.data[[patient.name]]$variable.regions[,c('chrom','arm','start.pos','end.pos')]
  if (is.null(regions) || nrow(regions) <= 0) next

@@

gene.cnt = ddply(var.regions, .(Gene.Symbols, CN), summarise, Progressors=length(which(status == 'P')), Non.Progressors=length(which(status == 'NP')))
gene.cnt[,unique(var.regions$patient.name)] = 0
gene.cnt = gene.cnt[which(!is.na(gene.cnt$Gene.Symbols)),]

@@

m = melt(gene.p[,c(1:4)], id.vars=c('Gene.Symbols', 'CN'))
m$Gene.Symbols = factor(m$Gene.Symbols, levels=unique(m$Gene.Symbols[c(nrow(m):1)]) ) # order
m$CN = factor(m$CN)

ggplot(m, aes(y=Gene.Symbols, x=variable)) + 
  geom_tile(aes(fill=value), color='white') + 
  scale_fill_gradient('n patients',low='white',high='red') + labs(x="",y='CGC Genes') +
  geom_point(data=m[m$CN == 'gain' & m$value > 0,], aes(y=Gene.Symbols, x=variable), size=2, shape=24, fill='red', color='red') +
  geom_point(data=m[m$CN == 'loss' & m$value > 0,], aes(y=Gene.Symbols, x=variable), color='green4', size=2, shape=25, fill='green4')

gc = ddply(gene.cnt, .(Gene.Symbols), summarise, Progressors=sum(Progressors), Non.Progressors=sum(Non.Progressors))
gc = arrange(subset(gc, Progressors >= Non.Progressors), -Progressors)
gc = gc[1:60,]
m = melt(gc, id.vars=c('Gene.Symbols'))
#m$value = as.factor(m$value)
m$Gene.Symbols = factor(m$Gene.Symbols, levels=unique(m$Gene.Symbols[c(nrow(m):1)]) ) # order
ggplot(m, aes(y=Gene.Symbols, x=variable)) + 
  geom_tile(aes(fill=value), color='white') + 
  scale_fill_gradient('n patients',low='white',high='red') + labs(x="",y='CGC Genes') + ggtitle("Top 50 genes only")

@@

pvnp = data.frame(matrix(nrow=length(subset(sum.patient.data, Status == 'P')$Patient), ncol=2, dimnames=list(subset(sum.patient.data, Status == 'P')$Patient, c('hgd','be'))))    
for (name in subset(sum.patient.data, Status == 'P')$Patient) {
  var.regions = patient.data[[name]]$variable.regions
  if ( is.null(var.regions) || nrow(var.regions) == 0 ) next

  sn = which(colnames(patient.data[[name]]$variable.regions)[-(1:5)] %in% subset(patient.data[[name]]$info, grepl('HGD|IMC', Pathology))$Samplename)
  if (length(sn) <= 0 | nrow(patient.data[[name]]$variable.regions) <= 0) next

@@

```{r, warning=F, eval=F}
#This is just looking at the distribution of values per sample. The average ploidy per sample is around 1.  Currently we're calling variable regions based on this.
ks = list()
for (pt in names(patient.data)) {
  print(pt)
  cols = intersect(arrange(patient.data[[pt]]$info, Pathology, Endoscopy.Year)$Samplename, grep('^D\\d+', colnames(patient.data[[pt]]$seg.vals), value=T))
  if (length(cols) <= 1) next

  segvals = patient.data[[pt]]$seg.vals[,cols]

### 2017-05-17 RegressionPvNP.Rmd
…

```{r apclust-func, echo=F, message=F, warning=F}
apclust.data<-function(segdata, samples) {
  x1 = segdata[,intersect(colnames(segdata), samples)]
  x1 = x1[, samples]
  rownames(x1) = (segdata[,c(1:4)] %>%
                    rowwise() %>%
                    mutate(location=paste(paste(chrom, arm, sep=''), '.', start.pos, '-', end.pos, sep='')))$location
  return(x1)  
}
```

@@

info = do.call(rbind, lapply(patient.data, function(df) df$info))
pg.samp = lapply(patient.data, function(pt) {
  info = pt$info
  info$SampleSD = apply(pt$seg.vals[,info$Samplename], 2, sd)
  info$SampleMEAN =  apply(pt$seg.vals[,info$Samplename], 2, mean)
  info$Prediction = NA
  info$Prediction.Dev.Resid = NA
  info$PID = unlist(lapply(info$Path.ID, function(x) unlist(strsplit(x, 'B'))[1]))
  return(info)
})

if (file.exists('/tmp/loo.Rdata')) {
  load('/tmp/loo.Rdata', verbose=T)
} else {
  performance.at.1se = c(); coefs = list(); plots = list(); fits = list()
  # Remove each patient (LOO)
  for (pt in names(pg.samp)) {
    tmp.patient.data = patient.data[subset(sum.patient.data, Patient != pt)$Patient]
    samples = as.vector(unlist(sapply(tmp.patient.data, function(df) df$info$Samplename )))
    length(samples)

    # Fit generated on all samples, including HGD
    a = 1
    fitLOO <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000) # all patients
    cv = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fitLOO$lambda, 
                                  pts=subset(pts, Samplename %in% samples), a=a, nfolds=10, splits=5, fit=fitLOO)
    #plots[[pt]] = cv$plot + ggtitle(pt)

@@

      pm = predict(fitLOO, dysplasia.df[pg.samp[[pt]]$Samplename,], s=cv$lambda.1se, type='response')
      sy = sqrt(binomial.deviance(pm, labels[pg.samp[[pt]]$Samplename]))

>>>>

```{r apclust-func, echo=F, message=F, warning=F}
apclust.data<-function(segdata, samples) {
  x1 = segdata[,intersect(colnames(segdata), samples)]

  if (length(samples) <= 1) { 
    x1 = as.data.frame(x1)
    colnames(x1) = samples
  } else {
    x1 = x1[, samples]
  }
  rownames(x1) = (segdata[,c(1:4)] %>%
                    rowwise() %>%
                    mutate(location=paste(paste(chrom, arm, sep=''), '.', start.pos, '-', end.pos, sep='')))$location
  return(x1)  
}
```

@@

info = do.call(rbind, lapply(patient.data, function(df) df$info))
pg.samp = lapply(patient.data, function(pt) {
  info = pt$info
  print(unique(info$Patient))
  info$SampleSD = NA
  info$SampleMEAN = NA
  if (length(info$Samplename) > 1) {
    info$SampleSD = apply(pt$seg.vals[,info$Samplename], 2, sd)
    info$SampleMEAN =  apply(pt$seg.vals[,info$Samplename], 2, mean)
  }
  info$Prediction = NA
  info$Prediction.Dev.Resid = NA
  info$PID = unlist(lapply(info$Path.ID, function(x) unlist(strsplit(x, 'B'))[1]))
  return(info)
})

if (file.exists('/tmp/loo.Rdata')) {
  load('/tmp/loo.Rdata', verbose=T)
} else {
  performance.at.1se = c(); coefs = list(); plots = list(); fits = list()
  # Remove each patient (LOO)
  for (pt in names(pg.samp)) {
    tmp.patient.data = patient.data[subset(sum.patient.data, Patient != pt)$Patient]
    samples = as.vector(unlist(sapply(tmp.patient.data, function(df) df$info$Samplename )))

    train.rows = which(rownames(dysplasia.df) %in% samples)
    training = dysplasia.df[train.rows,]
    test = dysplasia.df[-train.rows,]

    # Predict function giving me difficulty when I have only a single sample, this ensures the dimensions are the same
    sparsed_test_data <- Matrix(data=0, nrow=ifelse(nrow(test) > 1, nrow(test), 1),  ncol=ncol(training),
                            dimnames=list(pg.samp[[pt]]$Samplename,colnames(training)), sparse=T)
    for(i in colnames(test)) sparsed_test_data[,i] = test[,i]

    # Fit generated on all samples, including HGD
    a = 1
    fitLOO <- glmnet(training, labels[train.rows], alpha=a, family='binomial', nlambda=1000) # all patients
    cv = crossvalidate.by.patient(x=training, y=labels[train.rows], lambda=fitLOO$lambda, 
                                  pts=subset(pts, Samplename %in% samples), a=a, nfolds=10, splits=5, fit=fitLOO)
    #plots[[pt]] = cv$plot + ggtitle(pt)

@@

      pm = predict(fitLOO, newx=sparsed_test_data, s=cv$lambda.1se, type='response')
      sy = sqrt(binomial.deviance(pm, labels[pg.samp[[pt]]$Samplename]))
